Less Funding, More Lecturers, and Fewer Professors: \\ \vspace{0.1cm} Stagnating State Funding for Higher Education and its Effect on Faculty

### 2024-05-28 text/sections/04-empirical.tex
Finished adjusting the writing.  It is ready to Zenodo the data files.

This paper identifies the effects of changes in state funding on faculty.
However, a university's state funding is not exogenous to state decisions for support of higher education.
Instead, the state government undertakes a complex process of budgeting funding across many priorities, while higher education and faculty outcomes may not be independent in this process.
For example, it may be the case that a state government cuts funding for higher education as the demographics of the state change, and the average age grows older (such as in Maine) 
At the same time, it may be harder to attract younger faculty to work at universities in a state with an older population, so that faculty outcomes are correlated with state funding thanks to changes in state demographics.
This means that faculty outcomes would not be conditionally independent of faculty outcomes, and correlations between the two are not causal relationships.
Shifts in state for higher education, however, provide opportunity to address such endogeneity.

\subsection{State Funding Shift-Share}
\label{sec:approp-shocks}
I use a shift-share instrument to address endogeneity concerns for state governments' funding decisions, for their public universities.
\cite{NBERw23736,NBERw27885} develop the shift-share approach for public university funding by exploiting shifts in yearly state-wide higher eduction funding, interacted with universities' reliance on state funding in a base period.
\begin{align}
    \label{eqn:public-instrument}
    Z_{i,t} &\coloneqq - \left[
    \left( \frac{\text{Total State Funding}_{s(i),t}}{\text{Student Population}_{s(i),t}} \right)
    \sum_{\tau = 0}^{3} \frac 14
    \left( \frac{\text{State Funding}_{i,1990 + \tau}}{\text{Total Revenues}_{i,1990 + \tau}} \right) \right]
\end{align}

The system exploits the fact that institutions who rely more on state funding will be affected by follow-on state-wide funding cuts.
$Z_{i,t}$ is the shift-share instrument for institution $i$ in year $t$.
It is constructed by interacting the average funding for universities in state $s(i)$ with reliance on state funding relative to total revenues, averaged across the base years 1990--1993.\footnote{
    1990--1993 are defined as data for public university finance data are most comparable (i.e., without many missing values) beginning in 1990.
    \cite{NBERw23736} use the single year 1990 as the base year, though I use the four years to ameliorate missing values in the single year of 1990.
    Results are similar in either specification.
}
$Z_{i,t}$ is negative to reflect the fact that the long term trend in, and most of the short-run shifts in, state funding for higher education have been negative.\footnote{
    When used in log terms, the instrument is the negative of the logged shock --- i.e., $- \log \left( -Z_{i,t} \right)$ as opposed to log of the negative shock $\log Z_{i,t}$ directly.
}
State funding has been falling, so that the instrument describes shocks to university revenues, mostly in a negative direction.\footnote{
    \label{foot:control}
    \cite{NBERw27885} note the tendency for public universities to respond to state funding cuts by increasing reliance on tuition, where \cite{NBERw23736} specifically instruments for tuition revenues with collected information on legislative tuition price controls.
    It may be argued that tuition revenues are confounder between the causal effect of changes in state funding on a university's total revenues, so that this analysis focuses on state support for higher education (not total revenue), as do \cite{NBERw27885}.
    On the other hand, rises in tuition revenues (per student) may arise as result of tuition hikes thanks declining state support, which would mean controlling for tuition would constitute a bad control.
    Nonetheless, estimates including tuition revenues (per student) as a control in the second stage of the IV estimates produce results of very similar magnitude and direction, and so are omitted.
}

@@

The conditional independence assumption for the shift-share instrument makes sense in theory for state funding, but is fundamentally untestable with data.
However, the assumption is bolstered by the fact that universities with higher values of the shift-share instrument differ little in observed characteristics compared to those with lower values \citep{pei2019poorly}.
Universities are by-in-large smaller (in terms of enrolment, total revenues, and professor count) at the top of the shift-share instrument distribution (see~\autoref{tab:summary-quantiles}).
Though on a per student basis, there is little difference across the distribution (except in the outcomes under consideration).
The shift-share instrument is positively associated with the total enrolment and total amount of state funding for each university, in both \$ and $\log$/percentage change terms, while other the other sources of university finances are not associated with the funding shift-share.
So that the other sources of finances are not clear confounders for the IV strategy, as they exhibit balance with respect to the funding shift-share.
Nonetheless, yearly and institution fixed effects are included in regressions throughout to implicitly condition on mean differences between universities and years.

@@

Additionally, I investigate the effect of changes in state funding among incumbent professors using the Illinois data (IBHED).
Incumbent professors are faculty who are already employed at the university; state funding changes may affect faculty who are already at the university (beyond affecting whether they get hired), so I base the instrument in the year that the professor was hired, and include fixed effects for the hiring year.\footnote{
    This formulation follows that presented by \cite{NBERw27885}, where individual student outcomes are analysed via variation in state funding after their freshman-year.
    This contrasts with \autoref{sec:iv-model-uni} and \cite{NBERw23736}, where the unit of analysis is the university-year, where base year 1990 is more appropriate.
    See \autoref{sec:iv-model-indiv} for the instrument and second-stage specification.
}
The instrument exogeneity and exclusion follows the same argument as above, where no single Illinois campus takes the majority of state funding, and so
the Illinois state government's funding cuts are not majority aimed at any single campus.

It is not a priori clear which units are appropriate for this analysis;
does it make sense to consider state funding in purely dollar amounts per student, or in percent change rate?
The funding shift-share is a strong predictor for the average university's level of state funding in either unit; a funding shift-share of \$1,000 per student in the entire state leads to around \$1,176 per student at the university, while a funding shift-share of $-10$\% leads to around 9.77\% less state funding per year.
Yet, the level of state funding (and the outcome variables) vary greatly between states for the unit of analysis.
For example, the average Illinois public university receives \$10,709 in state funding per student in 1990 and  \$6,713 in 2021 (a fall of over 30\%), while California went from \$19,224 per student to \$12,915 in the same time span (a fall of 37\%).
While most states are not exactly the same as California or Illinois, this is example is telling for the phenomenon of stagnating state funding:
states vary in their absolute funding for higher education in 1990, but most have experienced a decline of around 30\% in per student funding, so that the trend is a common phenomenon in percentage terms across the US.
As such, I include regression specifications where the explanatory and outcome variables are $\log$ transformed, and refer to these when stating results in percent change terms.

\subsection{Effects in Years After the Initial Funding Cut}
\label{sec:local-projections}
The effects of a change in the universities funding may not be immediate, and faculty may be affected multiple years after a funding cut.
Yet, the funding shift-share instrument is significantly auto-correlated year-on-year; a large state-level funding cut in year $t$ is also likely to experience a large funding cut in year $t-1$.
\autoref{fig:lag-firststage} shows that the shift-share instrument is correlated with state funding five years into the past and future.
This level of time-series correlation would confound IV estimates of the effect of state funding in year $t$ on faculty outcomes in later years $t+k$, providing inconsistent estimates for the causal relationships.

>>>>

This paper identifies the effects of changes in state funding on faculty.
However, a university's state funding is not exogenous to state decisions for support of higher education.
Instead, the state government undertakes a complex process of budgeting funding across many priorities, while higher education and faculty outcomes may not be independent in this process.
This means that faculty outcomes would not be conditionally independent of faculty outcomes, and correlations between the two are not causal relationships.
Shifts in state for higher education, however, provide opportunity to address such endogeneity.

\subsection{State Funding Shift-Share}
\label{sec:approp-shocks}
I use a shift-share instrument variables (IV) approach to address endogeneity concerns for state governments' higher education funding decisions.
\cite{NBERw23736,NBERw27885} first developed the shift-share approach for public university funding by exploiting shifts in yearly state-wide higher eduction funding, interacted with universities' reliance on state funding in a base period.
$Z_{i,t}$ is the shift-share instrument for institution $i$ in year $t$:
\begin{align}
    \label{eqn:public-instrument}
    Z_{i,t} &\coloneqq - \left[
    \left( \frac{\text{Total State Funding}_{s(i),t}}{\text{Student Population}_{s(i),t}} \right)
    \sum_{\tau = 0}^{3} \frac 14
    \left( \frac{\text{State Funding}_{i,1990 + \tau}}{\text{Total Revenues}_{i,1990 + \tau}} \right) \right].
\end{align}

The system exploits the fact that institutions who rely more on state funding will be affected by follow-on state-wide funding cuts.
It is constructed by interacting the average funding for universities in state $s(i)$ with reliance on state funding relative to total revenues, averaged across the base years 1990--1993.\footnote{
    1990--1993 are defined as data for public university finance data are most comparable (i.e., without many missing values) beginning in 1990.
    \cite{NBERw23736} use the single year 1990 as the base year, though I use the four years to ameliorate missing values in the single year of 1990.
    Results are similar in either specification.
}
$Z_{i,t}$ is negative to reflect the fact that the long term trend in, and most of the short-run shifts in, state funding for higher education have been negative.\footnote{
    When used in log terms, the instrument is the negative of the logged shock --- i.e., $- \log \left( -Z_{i,t} \right)$ as opposed to log of the negative shock $\log Z_{i,t}$ directly.
}
State funding has been falling, so that the instrument describes shocks to university revenues, mostly in a negative direction.\footnote{
    \label{foot:control}
    \cite{NBERw27885} note the tendency for public universities to respond to state funding cuts by increasing reliance on tuition, where \cite{NBERw23736} specifically instruments for tuition revenues with collected information on legislative tuition price controls.
    It may be argued that tuition revenues are confounder between the causal effect of changes in state funding on a university's total revenues, so that this analysis focuses on state support for higher education (not total revenue), as do \cite{NBERw27885}.
    On the other hand, rises in tuition revenues (per-student) may arise as result of tuition hikes thanks declining state support, which would mean controlling for tuition would constitute a bad control.
    Nonetheless, estimates including tuition revenues (per-student) as a control in the second stage of the IV estimates produce results of very similar magnitude and direction, and so are omitted.
}

@@

The conditional independence assumption for the shift-share instrument makes sense in theory for state funding, but is fundamentally untestable with data.
However, the assumption is bolstered by the fact that universities with higher values of the shift-share instrument differ little in observed characteristics compared to those with lower values \citep{pei2019poorly}.
Universities are by-in-large smaller (in terms of enrolment, total revenues, and professor count) at the top of the shift-share instrument distribution (see~\autoref{tab:summary-quantiles}).
Though on a per-student basis, there is little difference across the distribution (except in the outcomes under consideration).
The shift-share instrument is positively associated with the total enrolment and total amount of state funding for each university, in both \$ and $\log$/percentage change terms, while other the other sources of university finances are not associated with the funding shift-share.
So that the other sources of finances are not clear confounders for the IV strategy, as they exhibit balance with respect to the funding shift-share.
Nonetheless, yearly and institution fixed effects are included in regressions throughout to implicitly control for differences between universities and years.

@@

Additionally, I investigate the effect of state funding cuts on professors using the Illinois data (IBHED).
State funding cuts may affect faculty in years after they join the university, so I base the instrument in the year that the professor was hired, and include fixed effects for the hiring year.\footnote{
    This formulation follows that presented by \cite{NBERw27885}, where individual student outcomes are analysed via variation in state funding after a student's freshman-year.
    This contrasts with \autoref{sec:iv-model-uni} and \cite{NBERw23736}, where the unit of analysis is the university-year, where base year 1990 is more appropriate.
    See \autoref{sec:iv-model-indiv} for the instrument and second-stage specification.
}
The instrument exogeneity and exclusion follows the same argument as above, where no single Illinois campus takes the majority of state funding, and so
the Illinois state government's funding cuts are not majority aimed at any single campus.

It is not a priori clear which units are appropriate for this analysis;
does it make sense to consider state funding in purely dollar amounts per-student, or in percent change rate?
The funding shift-share is a strong predictor for the average university's level of state funding in either unit; a funding shift-share of \$1,000 per-student in the entire state leads to around \$1,176 per-student at the university, while a funding shift-share of $-10$\% leads to around 9.77\% less state funding per year.
Yet, the level of state funding (and the outcome variables) vary greatly between states for the unit of analysis.
For example, the average Illinois public university receives \$10,709 in state funding per-student in 1990 and  \$6,713 in 2021 (a fall of over 30\%), while California went from \$19,224 per-student to \$12,915 in the same time span (a fall of 37\%).
While most states are not exactly the same as California or Illinois, this is example is telling for the phenomenon of stagnating state funding:
states vary in their absolute funding for higher education in 1990, but most have experienced a decline of around 30\% in per-student funding, so that the trend is a common phenomenon in percentage terms across the US.
As such, I include regression specifications where the explanatory and outcome variables are $\log$ transformed, and refer to these when stating results in percent change terms.

\subsection{Effects in Years After the Initial Funding Cut}
\label{sec:local-projections}
The effects of a change in the universities funding may not be immediate, and faculty may be affected multiple years after a funding cut.
Yet, the funding shift-share instrument is significantly auto-correlated year-on-year; a large state funding cut in year $t$ is also likely to experience a large funding cut in years $t \pm 1$.
\autoref{fig:lag-firststage} shows that the shift-share instrument is correlated with state funding five years into the past and future.
This level of time-series correlation would confound IV estimates of the effect of state funding in year $t$ on faculty outcomes in later years $t+k$, providing inconsistent estimates for the causal relationships when using two-stage least squares.

### 2024-05-28 text/sections/05-results.tex
Finished adjusting the writing.  It is ready to Zenodo the data files.

\subsection{First Stage Estimates}
The shift-share instrument has a strong first stage for universities' yearly state funding.
\autoref{tab:firststage-reg} presents results of the first-stage regression among IPEDS data, separately with and without a control for tuition revenue per student, plus institution and year fixed effects.
\autoref{tab:firststage-reg} Panel A shows estimates of a state funding shift-share of \$1 per student on state funding per student at the university, and Panel B the effects of a 1\% increase in the state funding shift-share on state funding per student.
Column (1) shows that a state funding shift-share of \$10 per student is associated with \$11.76 less state funding per student at the university; -10\% state funding shift-share leads to -9.77\% state funding per student at the average university.
Estimates are presented for specifications with and without fixed effects.
Column (2) shows estimates of the first-stage without including fixed effects, and gives less precise estimates for the funding shift-share --- this may be due to systematic differences in universities, which are not accounted for without fixed effects.
Columns (3) and (4) include the tuition revenue as a control (explained in \autoref{foot:control}) to exhibit estimates with or without this control variable.
Column (3) shows similar estimates to column (1) in both Panels A and B thanks to inclusion of fixed effects, so that 
whether tuition revenue is a bad control does not affect empirical estimates.
Column (1) represents the estimates for \autoref{eqn:firststage} with fixed effects (not including tuition revenue as a control variable), and is the preferred form that I proceed with.
I use the same empirical strategy in the analysis of Illinois faculty, and \autoref{tab:firststage-illinois} presents first-stage estimates among the IBHED data; results are similar to those calculated in IPEDS data.
These results, show the funding shift-share instrument strongly predicts university revenues in the first-stage estimation.

The first stage is strong in both IPEDS and IBHED data, and the argument for exogeneity comes from independence of the state funding mechanism to any individual public university campus (as in \autoref{sec:empirics}).
The IV approach also assumes that shifts in state-wide funding for higher education (the instrument) affect outcomes only via affecting state funding for the university (exclusion restriction).
This assumption is fundamentally untestable, but is bolstered by the lack of relationship with university characteristics.
\autoref{tab:summary-quantiles} shows the mean of various university variables, across the instrument distribution.
The university--years in the lowest 20\% of the instrument distribution have more research spending than those in the next 20\%; this is to be expected, as university spending is heavily related to state--wide changes in funding.
On the other hand, the measures of selectivity (acceptance rate, 6--year graduation rate) are not substantially different across the distribution of the instrument.
This allays concerns that state-wide shifts in state funding are targetted more towards less selective universities (a possible confounder), and supports use of the shift-share IV empirical design.

Do state funding cuts meaningfully affect universities?
While the shift-share performs well in predicting state funding (i.e., a strong instrument), it could be the case that universities are not affected.
For expample, universities can respond to funding cuts by charging higher tuition fees; it could be the case that this entirely off-sets funding cuts, leading to no effects on the university.
I estimate the IV model on IPEDS data for areas of university spending  (\autoref{tab:expenditures-shock-reg}).
This exercise gauges the effects of state funding on areas of university spending to guage whether the first stage is economically meaningful, and to see which areas of higher education operations are affected by state funding cuts.
Following a state funding cut, universities spend less on instruction, student services, operations, and grant aid.
Additionally, they charge higher tuition fees, though this rise is at most a quarter of the state funding cut, so does not anywhere near off-set state funding cuts.
These results are in line with previous work showing state funding cuts (or limits) reduce student-focused spending \citep{NBERw23736}, financial aid \citep{miller2022making}, and lead to increased tuition fees \citep{bound2019public}.

\subsection{University Level, IPEDS}
\label{sec:results-ipeds}
Funding cuts lead to measurable changes in the number of faculty at US public universities.
\autoref{tab:facultycount-shock-reg} presents OLS and IV estimates for the effect of a change in state funding on the number of faculty at the university, for each faculty position (lecturer, assistant and full professors, total).
The results are presented in two different units: Panel A shows how a \$1 increase in state funding changed the number of faculty per students, while Panel B shows how a 1\% increase in state funding led to a percent change in the number of faculty per students.

An extra \$1,000 in state funding per student leads to the average university employing 6 fewer lecturers, while effects for the other positions are not discernible from zero (\autoref{tab:facultycount-shock-reg}, Panel A).
A 10\% state funding cut lead to a 4.4\% increase in the number of lecturers per student, and a decrease of 1.35\% and 1.37\% for assistant and full professors, respectively.
A state funding cut of 10\% leads to a 0.65\% decrease in the total number of faculty per student (\autoref{tab:facultycount-shock-reg}, Panel B).
These are large substitution effects, and explain roughly 40\% of the trend in subsitutional towards lecturers over this same time period.
The average public university has roughly 36\% less state funding per student in 2021 than they did in 1990, while employing over double the number of lecturers and 20\% fewer professors (relative to enrolment, \autoref{fig:mean-funding-fte}, \ref{fig:fte-perprof}).

Once the funding cut has passed, universities do not re-adjust to hiring more professors to make up for last year's substitution towards lecturers.
\autoref{fig:count-lp} shows local projection estimates, where the estimate is the elasticity for professor count per student in year $t+k$ with respect to state funding in year $t$.
These elasticity estimates are shown in \autoref{tab:facultycount-shock-reg} Panel B, where the outcome is faculty count per students in later years.
State funding cuts affect the number of lecturers and professors at universities up to 9 years after the initial funding cut.
A 10\% state funding cut leads to 4.2\% (2.9--5.5 in a 95 percent confidence interval) more lecturers per student the following year, and around 3.1\% (0.6--5.5 confidence interval) more lecturers nine years later.
A 10\% state funding cut leads to 1.5\% (0.6--2.2 in a 95 percent confidence interval) fewer assistant professors per student the following year, and around 1.1\% (0.2--2.0 confidence interval) fewer nine years later;
the same cut leads to 1.2\% (0.6--1.8 in a 95 percent confidence interval) fewer full professors per student the following year, and around 0.9\% (0.1--1.7 confidence interval) fewer nine years later.

These effects are a substitution of professors for lecturers, as the total number of faculty per students are relatively unaffected in the follow-on years, while numbers of lecturers increase and professors decreases.
This agrees with a model for universities departments having fixed teaching requirements making human-resources decisions under constrained resources (e.g., \citealt{abe2015implications}), lending credibility to this type of model for understanding recent human resources decisions of public universities.
Furthermore, if a university must teach more students, but cannot afford to hire a full-time professor who teaches 2--3 classes a year, then it makes sense that they hire 2--3 additional lecturers on part-time contracts as a replacement.
This provides simple intuition for why the lecturers' elasticity is 2--3 times larger than that for assistant/full professors in \autoref{tab:facultycount-shock-reg}.
While no previous research causally connects funding cuts to substituting professors for lecturers, the phenomenon has been alluded to in popular media \citep{wiu2016} and university rankings (where more lecturers hurts a university's ranking, \citealt{usnews2023}); these results provides real evidence for these claims.

\subsection{Illinois Faculty, IBHED}
Incumbent professors at Illinois universities are relatively unaffected by state funding cuts for their university.
\autoref{tab:faculty-shock-illinois-rolling} show estimates of a 10\% change in state funding per student on faculty salaries, rate of exiting the Illinois public university system, and rate of promotion.
The estimates are not discernible from zero in any specification, including the shift-share IV estimates, so that faculty seem to have little or no state funding cuts passed on to them for these outcomes.

@@

These results show that faculty --- both lecturers and professors --- are relatively unaffected by state funding cuts, in the Illinois public university system.
Illinois is a state with a large public university system, relative to student enrolment, with trends in state funding for education similar to national trends, and with both selective and non-selective universities (\autoref{sec:trends-illinois}).
If faculty are unaffected in Illinois, then it is unlikely that faculty in the rest of the nation are affected on average.
Note, however, that this analysis studies faculty who are already faculty.
The IBHED data do not measure the entire pool of academics applying for faculty jobs at Illinois universities, but only those who applied to, were accepted, and then started a job as a lecturer or professor.
These are incumbent faculty, and it is clear they are unaffected by state funding cuts.
This raises a question: how can funding cuts lead to a large substitution of professors for lecturers, when incumbent faculty are not affected?

\subsection{Impacts on Faculty Hiring}
Turn-over is the primary channel that changes in faculty composition happen, where, where faculty exit their jobs and new faculty are hired to replace those leaving --- or no more are hired to replace those who left.
\autoref{fig:exit-illinois-lp-rolling} shows that Illinois faculty are no more likely to exit the Illinois system (either retire or take a new job elsewhere) following state funding cuts, so there is one remaining channel for the substitution of professors for lecturers: increased hiring of lecturers and reduced hiring of new professors.

@@

\autoref{tab:facultyhires-illinois-reg} shows these estimates for Illinois public universities.
In general, faculty counts are relatively unrelated to state funding among Illinois universities; though the point estimates line up with analysis in IPEDS, the smaller number of observations mean that the precision of estimates are small and indistinguishable from zero, in both raw counts and in log units.
On the other hand, universities with more state funding hire more professors across 2011--2021, as seen in \autoref{fig:hiring-correlation}.
An extra 10\% in state funding is associated with 8.5\% more professors hired per students.
I use these data as part of a shift-share analysis, showing that a 10\% state funding cut per student leads to 13\% fewer professor hires (relative to enrolment), and that these hires are roughly equal for gender of professor hired.

State funding cuts lead to lower professor hiring, and the persistent funding cuts to public universities led to lower professor hiring across the last three decades.
The channels for faculty turn-over were unaffected by state funding cuts in the analysis of Illinois faculty, and rate professor hiring is related to professor hiring in both correlational terms and in a shift-share analysis.
This means that effects on faculty hiring were likely the main mechanism for how state funding cuts led to a substitution of professors for lecturers at public universities.

\subsection{Robustness Checks}
\label{sec:results-robustness}
It is possible that the identification strategy is not robust to some confounding issues or trends.
For example, selective institutions may have received funding cuts for other reasons, or states with lower performing institutions or declining populations strategically cut funding for only less selective universities.
I address concerns for this with three different analyses.

@@

Secondly, I return to the primary specification in \autoref{eqn:secondstage}, and estimate effect of state funding cuts on faculty counts separately among universities of different selectivity rates.
This investigates whether results are consistent across different levels of selectivity, and whether less selective universities have unreasonably high estimates for substitution --- as would be the case if selectivity was an unaddressed confounder.
\autoref{tab:facultycount-heterogeneity} collects these results together, separated by the \cite{barrons2009} selectivity rankings.
Barrons rank 1-2 are the most selective (including e.g., UC Berkeley and SUNY Binghamton), 3-4 selective (including e.g., Clemson University and University of Dallas), and rank 5 for the rest of universities.
The first-stage is strong in every group of universities, even among the small sample size of the 12 universities in the most selective group, and the 28 universities in the next selective group.
Faculty substitution rates are not very strong among the most selective universities, but are very similar between selective universities and unranked universities, showing that even selective public universities exhibit the substitution towards lecturers.

>>>>

\subsection{First Stage Estimates}
The shift-share instrument has a strong first stage for universities' yearly state funding.
\autoref{tab:firststage-reg} show results of the first-stage regression among IPEDS data, separately with and without a control for tuition revenue per student.
\autoref{tab:firststage-reg} Panel A shows estimates of a state funding shift-share of \$1 per student on state funding per student at the university, and Panel B the effects of a 1\% increase in the state funding shift-share on state funding per student.
Column (1) shows that a state funding shift-share of \$10 per student is associated with \$11.76 less state funding per student at the university; $-10$\% state funding shift-share leads to $-9.77$\% state funding per student at the average university.
Estimates are presented for specifications with and without fixed effects.
Column (2) shows estimates of the first-stage without including fixed effects, and gives less precise estimates for the funding shift-share --- this may be due to systematic differences in universities, which are not accounted for without fixed effects.
Columns (3) and (4) include the tuition revenue as a control (explained in \autoref{foot:control}) to exhibit estimates with or without this control variable.
Column (3) shows similar estimates to column (1) in both Panels A and B thanks to inclusion of fixed effects, so whether tuition revenue is a bad control does not affect empirical estimates.
Column (1) represents the estimates for \autoref{eqn:firststage} with fixed effects (not including tuition revenue as a control variable), and is the preferred specification that I proceed with.
I use the same empirical strategy in the analysis of Illinois faculty, and \autoref{tab:firststage-illinois} presents first-stage estimates among the IBHED data; results are similar to those calculated in IPEDS data.
These results show the funding shift-share instrument strongly predicts a university's state funding in the first-stage estimation.

The first stage is strong in both IPEDS and IBHED data, and the argument for exogeneity comes from independence of the state funding mechanism to any individual public university campus (as in \autoref{sec:empirics}).
The IV approach also assumes that shifts in state-wide funding for higher education (the instrument) affect outcomes only via affecting state funding for the university (exclusion restriction).
This assumption is fundamentally untestable, but is bolstered by the lack of relationship with university characteristics.
\autoref{tab:summary-quantiles} shows the mean of various university variables, across the instrument distribution.
The university--years in the bottom 20\% of the instrument distribution have more research spending than those in the next 20\%; this is to be expected, as university spending is heavily related to state--wide shifts in funding.
On the other hand, the measures of selectivity (acceptance rate, 6--year graduation rate) are not substantially different across the distribution of the instrument.
This allays concerns that state-wide shifts in state funding are targetted at less selective universities (a possible confounder), and supports use of the shift-share IV empirical design.

Do state funding cuts meaningfully affect universities?
While the shift-share performs well in predicting state funding (i.e., a strong instrument), it could be the case that the day-to-day operations of a university are not affected.
University could have responded to state funding cuts by cutting areas of spending unrelated to research or instruction, or fully off-set the cuts by charging higher tuition fees, leading to no effects on the university or faculty.
To allay this concern, I estimate the IV model on IPEDS data for areas of university spending  (\autoref{tab:expenditures-shock-reg}).
This exercise gauges the effects of state funding cuts on areas of university spending to gauge whether the first stage is economically meaningful, and to see which areas of higher education operations are affected by state funding cuts.
Following a state funding cut, universities spend less on instruction, student services, operations, and grant aid.
Additionally, they charge higher tuition fees, though this rise is at most a quarter of the state funding cut, so does not come close to fully off-setting state funding cuts.
These results are in line with previous work showing state funding cuts (or tuition price limits) reduce student-focused spending \citep{NBERw23736}, financial aid \citep{miller2022making}, and lead to increased tuition fees \citep{bound2019public}.

\subsection{University Level, IPEDS}
\label{sec:results-ipeds}
State funding cuts lead to measurable changes in the number of faculty at US public universities.
\autoref{tab:facultycount-shock-reg} presents OLS and IV estimates for the effect of state funding cuts on the number of faculty at public universities, separately for each faculty position (lecturer, assistant and full professors, total).
The results are presented in two different units: Panel A shows how a \$1 increase in state funding changed the count of faculty per students, while Panel B shows how a 1\% increase in state funding led to a percent change in the number of faculty per students.

An extra \$1,000 in state funding per student leads to the average university employing 6 fewer lecturers, while effects for the other positions are not discernible from zero (\autoref{tab:facultycount-shock-reg}, Panel A).
A 10\% state funding cut lead to a 4.4\% increase in the number of lecturers per student, and a decrease of 1.35\% and 1.37\% for assistant and full professors, respectively.
A state funding cut of 10\% leads to a 0.65\% decrease in the total number of faculty per student (\autoref{tab:facultycount-shock-reg}, Panel B).
These are large substitution effects, and explain roughly 40\% of the trend in substitution towards lecturers over the same time period.
The average public university has roughly 36\% less state funding per student in 2021 than they did in 1990, while employing over double the number of lecturers and 20\% fewer professors (relative to enrolment, seen in \autoref{fig:mean-funding-fte}, \ref{fig:fte-perprof}).

Universities do not re-adjust by hiring more professors to make up for last year's funding cut, and so the substitution towards lecturers becomes persistent.
\autoref{fig:count-lp} shows local projection estimates, showing the elasticity for professor count per student in year $t+k$ with respect to state funding in year $t$.
State funding cuts affect the number of lecturers and professors at universities up to 9 years after the initial funding cut.
A 10\% state funding cut leads to 4.2\% (2.9--5.5 in a 95 percent confidence interval) more lecturers per student the following year, and around 3.1\% (0.6--5.5 confidence interval) more lecturers nine years later.
A 10\% state funding cut leads to 1.5\% (0.6--2.2 in a 95 percent confidence interval) fewer assistant professors per student the following year, and around 1.1\% (0.2--2.0 confidence interval) fewer nine years later;
the same cut leads to 1.2\% (0.6--1.8 in a 95 percent confidence interval) fewer full professors per student the following year, and around 0.9\% (0.1--1.7 confidence interval) fewer nine years later.

These effects are a substitution of professors for lecturers at public universities, as the total number of faculty per students are relatively unrelated to funding cuts, but the lecturer count increases and professor count decreases.
This agrees with a model for universities departments having fixed teaching requirements making human-resources decisions under constrained resources (e.g., \citealt{abe2015implications}), lending credibility to this type of model for understanding recent human resources decisions of public universities.
Furthermore, if a university must teach more students, but cannot afford to hire a full-time professor who teaches 2--3 classes a year, then it makes sense that they hire 2--3 additional lecturers on part-time contracts for each class as a replacement.
This provides simple intuition for why the lecturers' elasticity is 2--3 times larger than that for assistant/full professors in \autoref{tab:facultycount-shock-reg}.
While no previous research causally connects funding cuts to substituting professors for lecturers, the phenomenon has been alluded to in popular media \citep{wiu2016} and university rankings (where more lecturers hurts a university's ranking, \citealt{usnews2023}).
The results provided here give the first real evidence for such popular claims.

\subsection{Illinois Faculty, IBHED}
Incumbent professors at Illinois universities are relatively unaffected by state funding cuts at their university.
\autoref{tab:faculty-shock-illinois-rolling} show estimates of a 10\% change in state funding per student on faculty salaries, rate of exiting the Illinois public university system, and rate of promotion.
The estimates are not discernible from zero in any specification, including the shift-share IV estimates, so that faculty seem to have little or no state funding cuts passed on to them for these outcomes.

@@

These results show that faculty --- both lecturers and professors --- are relatively unaffected by state funding cuts, in the Illinois public university system.
Illinois is a state with a large public university system, relative to student enrolment, with trends in state funding for education similar to national trends, and with both selective and non-selective universities.
If faculty are unaffected in Illinois, then it is unlikely that faculty in the rest of the nation are affected on average.
Note, however, that this analysis studies faculty who are already faculty.
The IBHED data do not measure the entire pool of academics applying for faculty jobs at Illinois universities, but only those who applied to, were accepted, and then started a job as a lecturer or professor.
These are incumbent faculty, and the data show they are unaffected by state funding cuts.
This raises a question: how can funding cuts lead to a large substitution of professors for lecturers, when incumbent faculty are not affected?

\subsection{Impacts on Faculty Hiring}
Turn-over is the primary channel that changes in faculty composition happen, where faculty exit their jobs and new faculty are hired to replace those leaving --- or no more are hired to replace those who left.
\autoref{fig:exit-illinois-lp-rolling} shows that Illinois faculty are no more likely to exit the Illinois system (either retire or take a new job elsewhere) following state funding cuts, so there is one remaining channel for the substitution of professors for lecturers: increased hiring of lecturers and reduced hiring of new professors.

@@

\autoref{tab:facultyhires-illinois-reg} shows these estimates for Illinois public universities.
In general, faculty counts are relatively unrelated to state funding among Illinois universities; though the point estimates line up with analysis in IPEDS, the smaller number of observations mean that the precision of estimates are small and indistinguishable from zero, in both raw counts and in log units.
In the national-level \cite{wapman2022quantifying} data, universities with more state funding hire more professors across 2011--2021.
An extra 10\% in state funding is associated with 8.5\% more professors hired per students, as seen in \autoref{fig:hiring-correlation}.
I use these data as part of a shift-share analysis, showing that a 10\% state funding cut per student leads to 13\% fewer professor hires (relative to enrolment), and that these hires are roughly equal for gender of professor hired.

State funding cuts lead to lower professor hiring, and the persistent funding cuts to public universities led to lower professor hiring across the last three decades.
The channels for faculty turn-over were unaffected by state funding cuts in the analysis of Illinois faculty, while professor hiring is related to funding cuts in both correlational terms and in a shift-share analysis.
Combining these pieces of evidence means that reduced professor hiring was likely the main mechanism for how state funding cuts led to a substitution of professors for lecturers at public universities.

\subsection{Robustness Checks}
\label{sec:results-robustness}
It is possible that the identification strategy is not robust to some confounding issues or trends.
For example, selective institutions may have received funding cuts for other reasons, or states with lower performing institutions or declining populations strategically cut funding for only less selective universities.
I address these concerns with three different analyses.

@@

Secondly, I return to the primary specification in \autoref{eqn:secondstage}, and estimate effect of state funding cuts on faculty counts separately among universities of different selectivity rates.
This investigates whether results are consistent across different levels of selectivity, and whether less selective universities have unreasonably high estimates for substitution --- as would be the case if selectivity were an unaddressed confounder.
\autoref{tab:facultycount-heterogeneity} collects these results together, separated by the \cite{barrons2009} selectivity rankings.
Barrons rank 1-2 are the most selective (including e.g., UC Berkeley and SUNY Binghamton), 3-4 selective (including e.g., Clemson University and University of Dallas), and rank 5 for the rest of universities.
The first-stage is strong in every group of universities, even among the small sample size of the 12 universities in the most selective group, and the 28 universities in the next selective group.
Faculty substitution rates are not very strong among the most selective universities, but are very similar between selective universities and unranked universities, showing that even selective public universities exhibit the substitution towards lecturers.
